Temporal and Spatial Analysis of Hate Crimes in Toronto

### 2024-09-27 paper/paper.qmd
Final committ

## Measurement and packages 
The Data set used in the analysis is gathered from Open Data Toronto through the Open Data Toronto [@opendatatoronto] and used the statistical software R [@citeR] for importing data, data cleaning and testing data. The data source "Crime Data" is collected from Toronto Police starting from 2018 to the end of 2023. It includes both temporal and spatial attributes. Other packages were used for analyzing the data such as ggplot [@ggplot2], knitr[@knitr], tidyverse [@tidyverse], lubridate [@lubridate], tinytex[@tinytex], dplyr [@dplyr].

@@

## Observation of cleaned data
A sample of observation of cleaned data is displayed in @tbl-1. This table includes key variables of interest that will be used for further analysis. Specifically, it contains the bias type, which indicates whether the crime was motivated by factors such as race, religion, or sexual orientation. The date of the incident is provided, allowing for temporal analysis of hate crime trends over time. Additionally, the neighborhood where each crime occurred is included, which is crucial for understanding geographic characteristics and identifying potential hotspots for hate crimes within Toronto. The offense type records the legal classification of the crime, such as assault or mischief providing insight into the nature of the incidents. Lastly, the arrest status variable indicates whether law enforcement made an arrest in response to the reported crime, which is an important indicator of justice system responses to hate crimes.

@@

A general summary statistic is shown @tbl-2. Since hate crimes are relatively rare events, the data may show an uneven distribution across different years and neighborhoods. Some years may exhibit higher counts due to specific incidents or socio-political factors that led to spikes in reporting, while other years may show fewer cases, resulting in an unbalanced distribution. Due to the categorical nature of many key variables, such as bias types (race, religion, sexual orientation), and the relatively small number of observations per category, more advanced summary statistics (e.g., mean, standard deviation) are not as meaningful in this context. Instead, the focus is on counting occurrences and understanding the distribution of incidents across time and location. Therefore, we limit the analysis to counts and proportions rather than providing more traditional numeric summary statistics. Further analysis of location and its relationship to number of hate crimes will be shown in @sec-results.

@@

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

## Measurement and Packages Used
The Data set used in the analysis is gathered from Open Data Toronto through the Open Data Toronto [@opendatatoronto] and used the statistical software R [@citeR] for importing data, data cleaning and testing data. The data source "Crime Data" is collected from Toronto Police starting from 2018 to the end of 2023. It includes both temporal and spatial attributes. Other packages were used for analyzing the data such as ggplot [@ggplot2], knitr[@knitr], tidyverse [@tidyverse], lubridate [@lubridate], tinytex[@tinytex], dplyr [@dplyr].

@@

## Observation of Cleaned Data
A sample of observation of cleaned data is displayed in @tbl-1. This table includes key variables of interest that will be used for further analysis. Specifically, it contains the bias type, which indicates whether the crime was motivated by factors such as race, religion, or sexual orientation. The date of the incident is provided, allowing for temporal analysis of hate crime trends over time. Additionally, the neighborhood where each crime occurred is included, which is crucial for understanding geographic characteristics and identifying potential hotspots for hate crimes within Toronto. The offense type records the legal classification of the crime, such as assault or mischief providing insight into the nature of the incidents. Lastly, the arrest status variable indicates whether law enforcement made an arrest in response to the reported crime, which is an important indicator of justice system responses to hate crimes.

@@

A general summary statistic is shown @tbl-2. Since hate crimes are relatively rare events, the data may show an uneven distribution across different years and neighborhoods. After inspecting the data age and language biases are dropped due to small sample size which indicates these two biases are less likely to be taken place in Toronto Area. Some years may exhibit higher counts due to specific incidents or political factors that led to spikes in reporting while other years may show fewer cases resulting in an unbalanced distribution. Due to the categorical nature of many key variables such as bias types (race, religion, sexual orientation), and the relatively small number of observations per category, more advanced summary statistics (e.g., mean, standard deviation) are not as meaningful in this context. Instead, the focus is on counting occurrences and understanding the distribution of incidents across time and location. Therefore, we limit the analysis to counts and proportions rather than providing more traditional numeric summary statistics. Further analysis of location and its relationship to number of hate crimes will be shown in @sec-results.

@@

\newpage
# Discussion {#sec-discussion}

## Rising Trend of Hate Crimes

The time-series analysis in @fig-1 reveals a steady increase in hate crimes in Toronto from 2018 to 2023 with spikes in 2020 and 2022. These years coincide with significant societal disruptions caused by the COVID-19 pandemic which drastically altered daily life, the economy and social dynamics worldwide. The pandemic not only led to a global health crisis but also intensified economic disparities, social isolation, and general uncertainty.

The pandemic’s economic impacts including layoffs from work and rising inflation intensified societal stresses. In @tbl-2 the basic summary statistics shows how hate crimes dramatically increase in 2020 which coincides with the time series graph in @fig-1 and the discussion in COVID-19 impacts. As many people faced severe financial problem and emotional strain marginalized communities often became scapegoats for these frustrations. In particular, black race communities worldwide experienced increased hate crimes. Further analysis on hate crimes towards black race is included in 
[Disproportionate Racial Bias](#racial-bias). This spike in hate crimes during 2020 and 2022 aligns with a broader pattern of rising xenophobia and racial targeting exacerbated by misinformation and heightened racial tensions. The pandemic appears to have not only exposed underlying societal inequalities but also amplified them.

Moreover, the upward trajectory of hate crimes during these years suggests that existing hate crime prevention policies may have been insufficient in addressing the root causes of such bias-motivated offenses. The stress brought by COVID-19 acted as a trigger for already existing tensions which implies that long-standing systemic issues regarding racial discrimination were not adequately addressed.

## Disproportionate Racial Bias {#racial-bias}
According to @fig-3, the discrimination towards black race individuals has a  significantly lead comparing to hate crimes towards other races. The data consistently show that hate crimes motivated by racial bias particularly those targeting Black individuals form the largest portion of the incidents across all years. This finding aligns with global concerns over racial discrimination and reinforces the idea that racial bias remains a significant driver of hate crimes in Toronto. Specific neighborhoods, such as Yonge-Bay Corridor and York University Heights show a high concentration of such incidents emphasizing the need for localized responses. These neighborhoods may require targeted community outreach and police presence to address racial tensions and foster exclusivity. The dominant status of racial bias in hate crimes suggests that efforts to combat racism must remain a central focus of anti-hate crime policies.

## Spatial Variances and Neighborhood Hotspots

The spatial analysis underscores that hate crimes are not evenly distributed across Toronto, with certain neighborhoods experiencing disproportionately higher rates of incidents shown in @tbl-3. For instance, the Yonge-Bay Corridor and Church-Wellesley neighborhoods consistently rank among the highest in hate crime occurrences. This geographic concentration suggests that specific local tensions may contribute to the higher rates of hate crimes in these areas. Furthermore in @tbl-4, the Yonge-Bay Corridor and Church-Wellesley neighborhoods continue to be prominent hotspots for hate crimes targeting Black individuals. This trend emphasizes the persistent racial bias and discrimination in these parts of the city reflecting broader societal challenges that Black communities face.These areas have seen a disproportionately high number of incidents compared to other neighborhoods in Toronto.  Understanding the root causes of these differences whether they stem from economic inequality, social marginalization or historical prejudices can guide more effective policy interventions. Targeted strategies that address the unique conditions of these neighborhoods are crucial in reducing the incidence of hate crimes and fostering safer and more inclusive communities.

## Weaknesses and next steps
The analysis of hate crimes in Toronto reveals several important findings but there are also notable limitations that must be addressed. First, the data used is limited to reported incidents which may not capture the full scope of hate crimes occurring in the city as many incidents go unreported due to fear distrust of authorities, or other reasons. Additionally, the analysis relies heavily on the available police data which may have biases in how crimes are classified or reported.
One significant limitation of the data structure is that it heavily relies on categorical variables such as race bias, religion bias, and sexual orientation bias without providing in-depth contextual information about each incident. This structure limits the ability to explore nuances within hate crimes such as intersecting identities (e.g., race and religion) or the background of the victims. Furthermore, the data does not account for the severity or outcome of each incident which could offer insight into the effectiveness of law enforcement responses. Another limitation is the time-frame of the data which still miss long-term trends or cyclical patterns in hate crimes.These limitations suggest that future research should aim to incorporate community surveys or qualitative studies to capture a broader perspective of hate crime experiences. Moving forward, a deeper exploration of the factors driving the concentration of hate crimes in specific neighborhoods and the effectiveness of existing interventions is crucial.
